---
title: "Breaking Ball Taxonomy: Curveball, Slider, and Sweeper"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 9
    fig_height: 5.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)

```

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(mclust)
library(jagsUI)
library(httr)
library(readr)
library(knitr)
library(tibble)
library(vegan)
library(ggrepel)
```



## Data

```{r data-pull}
breaking <- read_csv("C:/Users/fishd/OneDrive/Manuscripts/Statcast/WD/statcast_breaking.csv")

```

```{r pitch-counts-table}
count_table <- breaking |>
  dplyr::count(game_year, pitch_type) |>
  tidyr::pivot_wider(names_from = pitch_type, values_from = n, values_fill = 0)

kable(count_table,
      col.names   = c("Season", "CU", "SL", "ST"),
      caption     = "Pitch counts by type and season (CU = curveball, SL = slider, ST = sweeper).",
      format.args = list(big.mark = ","))
```



```{r data-prep}
REQUIRED_COLS <- c("pitcher", "game_year", "pitch_type", "p_throws",
                   "pfx_x", "pfx_z", "release_speed",
                   "release_spin_rate", "description")

breaking <- breaking |>
  dplyr::filter(pitch_type %in% c("CU", "SL", "ST")) |>
  select(any_of(REQUIRED_COLS)) |>
  mutate(
    pfx_x_adj = if_else(p_throws == "L", -pfx_x, pfx_x),
    whiff     = as.integer(description == "swinging_strike"),
    label_sl  = as.integer(pitch_type == "SL"),
    label_st  = as.integer(pitch_type == "ST"),
    pitcher_id = as.integer(factor(pitcher)),
    year_f     = factor(game_year)
  ) |>
  drop_na(pfx_x_adj, pfx_z, release_speed, release_spin_rate) |>
  mutate(
    pfx_x_std = as.numeric(scale(pfx_x_adj)),
    pfx_z_std = as.numeric(scale(pfx_z)),
    velo_std  = as.numeric(scale(release_speed)),
    spin_std  = as.numeric(scale(release_spin_rate))
  ) |>
  # Residualise year out of each feature so PCA reflects within-year
  # movement structure rather than secular between-year trends.
  mutate(
    pfx_x_std = residuals(lm(pfx_x_std ~ year_f, data = pick(everything()))),
    pfx_z_std = residuals(lm(pfx_z_std ~ year_f, data = pick(everything()))),
    velo_std  = residuals(lm(velo_std  ~ year_f, data = pick(everything()))),
    spin_std  = residuals(lm(spin_std  ~ year_f, data = pick(everything())))
  )

cat(sprintf(
  "%s pitches kept — CU: %s  SL: %s  ST: %s  |  %s unique pitchers\n",
  format(nrow(breaking),                        big.mark = ","),
  format(sum(breaking$pitch_type == "CU"),      big.mark = ","),
  format(sum(breaking$pitch_type == "SL"),      big.mark = ","),
  format(sum(breaking$pitch_type == "ST"),      big.mark = ","),
  format(n_distinct(breaking$pitcher),          big.mark = ",")
))

```

---

## Stage 1: PCA

PCA is run on four standardised features — horizontal break (`pfx_x_adj`, flipped
for LHP so glove-side movement is consistently negative), vertical break (`pfx_z`),
velocity, and spin rate — to characterise the shape of breaking-ball movement space.
The **continuum hypothesis** predicts a single elongated cloud in PC space rather
than three distinct clusters. This stage addresses **convergent validity**: do the
three Statcast labels occupy separable regions of physical movement space?

```{r pca-fit}
features_scaled <- breaking |>
  select(pfx_x_std, pfx_z_std, velo_std, spin_std) |>
  as.matrix()

rda_fit  <- rda(features_scaled)
rda_smry <- summary(rda_fit)

# Site scores on default vegan scale (gives sensible axis ranges)
all_scores <- as.data.frame(scores(rda_fit, choices = 1:2)$sites)
all_scores$pitch_type <- breaking$pitch_type

# Species scores scaled to fit inside the point cloud
species_raw <- as.data.frame(scores(rda_fit, choices = 1:2)$species)
species_raw$feature <- rownames(species_raw)

site_range  <- max(abs(c(all_scores$PC1, all_scores$PC2)))
spec_range  <- max(abs(c(species_raw$PC1, species_raw$PC2)))
arrow_scale <- 1 * site_range / spec_range

loadings_raw <- species_raw |>
  mutate(
    PC1  = PC1 * arrow_scale,
    PC2  = PC2 * arrow_scale,
    dist = sqrt(PC1^2 + PC2^2),
    label = recode(feature,
      pfx_x_std = "H. break",
      pfx_z_std = "V. break",
      velo_std  = "Velocity",
      spin_std  = "Spin rate"
    )
  )

# Variance explained
var_exp <- rda_smry$cont$importance["Proportion Explained", ]
cum_var <- rda_smry$cont$importance["Cumulative Proportion",  ]
```



```{r pca-variance-table}
var_tbl <- data.frame(
  PC         = paste0("PC", 1:4),
  Proportion = round(var_exp, 3),
  Cumulative = round(cum_var, 3),
  check.names = FALSE
)
kable(var_tbl, caption = "Variance explained by each principal component.")

```

```{r pca-loadings-table}
ldg <- loadings_raw |>
  select(Feature = feature, PC1, PC2) |>
  mutate(
    Feature = recode(Feature,
      pfx_x_std = "Horiz. break (pfx_x, adj.)",
      pfx_z_std = "Vert. break (pfx_z)",
      velo_std  = "Release speed",
      spin_std  = "Spin rate"
    ),
    across(where(is.numeric), ~ round(., 3))
  )
kable(ldg, caption = "PCA loadings (vegan rda).")
```



PC1 captures the horizontal/vertical break trade-off: curveballs load heavily on
vertical drop; sweepers load heavily on horizontal break. PC2 broadly separates
velocity and spin from movement magnitude.

```{r pca-scatter, fig.cap="PC1 vs PC2 coloured by Statcast pitch label. Convex hulls outline each label's occupied region; broadly overlapping regions support the continuum hypothesis. Arrows show feature loadings scaled to the point-cloud range."}
set.seed(1)
label_colours <- c(CU = "#0072B2", SL = "#D55E00", ST = "#009E73")
label_names   <- c(CU = "Curveball", SL = "Slider",  ST = "Sweeper")

scores_plot <- all_scores |> slice_sample(n = min(nrow(all_scores), 25000))

hull_df <- scores_plot |>
  group_by(pitch_type) |>
  slice(chull(PC1, PC2)) |>
  ungroup()

ggplot(scores_plot, aes(x = PC1, y = PC2)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_polygon(
    data = hull_df,
    aes(group = pitch_type, colour = pitch_type, fill = pitch_type),
    alpha = 0.15
  ) +
  geom_point(aes(colour = pitch_type), alpha = 0.15, size = 0.5) +
  geom_segment(
    data = loadings_raw,
    aes(x = 0, xend = PC1, y = 0, yend = PC2),
    colour = "black", linewidth = 0.65, alpha = 0.55,
    arrow = arrow(length = unit(0.01, "npc")),
    inherit.aes = FALSE
  ) +
  geom_label_repel(
    data = loadings_raw,
    aes(x = PC1, y = PC2, label = label),
    size = 4.35, alpha = 0.8,
    label.padding = unit(0.35, "lines"),
    inherit.aes = FALSE
  ) +
  scale_colour_manual(values = label_colours, labels = label_names) +
  scale_fill_manual(values   = label_colours, labels = label_names) +
  labs(
    x      = sprintf("PC1 (%.1f%% variance)", var_exp[1] * 100),
    y      = sprintf("PC2 (%.1f%% variance)", var_exp[2] * 100),
    colour = NULL, fill = NULL,
    title  = "PCA of breaking-ball movement space (2022–2025)"
  ) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "bottom") +
  guides(
    colour = guide_legend(override.aes = list(alpha = 1, size = 2)),
    fill   = guide_legend(override.aes = list(alpha = 0.4))
  )

```

---

## Stage 2: Gaussian Mixture Modelling

GMM via `mclust` is applied to the same four scaled features. The number of
components $G$ and the covariance parameterisation are selected jointly by BIC,
with no assumption that $G = 3$. Cluster–label alignment is then measured with
the Adjusted Rand Index (ARI). This stage addresses **discriminant validity**:
does a purely data-driven partition recover MLB's operational taxonomy?

Soft posterior assignment probabilities are retained to assess classification
certainty. If many pitches carry diffuse probabilities across components, that
is itself evidence against hard taxonomic boundaries.

```{r gmm-fit, cache=TRUE}
set.seed(42)
bic_obj <- mclustBIC(features_scaled, G = 2:9, verbose = FALSE)
fit_gmm <- Mclust(features_scaled, x = bic_obj, verbose = FALSE)

cat(sprintf("BIC-optimal: G = %d, model = %s\n",
            fit_gmm$G, fit_gmm$modelName))
```



```{r gmm-bic-curve, fig.cap="BIC values across candidate GMM components. Each line is a covariance parameterisation; the peak marks the preferred solution. A BIC-optimal G greater than three suggests MLB's taxonomy is a coarse discretisation of richer latent structure."}
bic_df <- as.data.frame(bic_obj) |>
  rownames_to_column("G") |>
  mutate(G = as.integer(G)) |>
  pivot_longer(-G, names_to = "model", values_to = "BIC") |>
  filter(!is.na(BIC))

bic_best <- bic_df |>
  group_by(G) |>
  slice_max(BIC, n = 1, with_ties = FALSE)

ggplot(bic_df, aes(G, BIC, group = model, colour = model)) +
  geom_line(alpha = 0.4, linewidth = 0.5) +
  geom_point(data = bic_best, aes(G, BIC), colour = "black", size = 2.5) +
  geom_vline(xintercept = fit_gmm$G, linetype = "dashed", colour = "red") +
  scale_x_continuous(breaks = 2:9) +
  labs(
    x      = "Number of GMM components (G)",
    y      = "BIC",
    colour = "Model",
    title  = sprintf("GMM BIC curve — optimal G = %d (%s)",
                     fit_gmm$G, fit_gmm$modelName)
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "right",
        legend.text      = element_text(size = 8))

```

```{r gmm-cluster-scatter, fig.cap="GMM cluster assignments overlaid on PC space, coloured by Statcast label. Clusters dominated by a single colour indicate label–cluster alignment; mixed-colour clusters indicate boundary ambiguity."}
set.seed(1)
cluster_df <- data.frame(
  PC1        = all_scores$PC1,
  PC2        = all_scores$PC2,
  pitch_type = breaking$pitch_type,
  cluster    = factor(fit_gmm$classification)
) |>
  slice_sample(n = min(nrow(all_scores), 25000))

ggplot(cluster_df, aes(PC1, PC2)) +
  geom_point(aes(colour = pitch_type, shape = cluster),
             alpha = 0.2, size = 0.7) +
  scale_colour_manual(values = label_colours, labels = label_names) +
  scale_shape_manual(values = c(16, 17, 15, 3, 4, 8, 7, 13)[seq_len(fit_gmm$G)]) +
  labs(
    x      = sprintf("PC1 (%.1f%%)", var_exp[1] * 100),
    y      = sprintf("PC2 (%.1f%%)", var_exp[2] * 100),
    colour = "Statcast label",
    shape  = "GMM cluster",
    title  = sprintf("GMM clusters (G = %d) vs. Statcast labels", fit_gmm$G)
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  guides(
    colour = guide_legend(override.aes = list(alpha = 1, size = 2)),
    shape  = guide_legend(override.aes = list(alpha = 1, size = 2))
  )
```



```{r gmm-uncertainty, fig.cap="Distribution of posterior classification uncertainty (1 minus the maximum component probability) by Statcast label. Pitches near zero are assigned with high confidence; pitches near 0.5 are genuinely ambiguous. A long right tail or a mode away from zero is consistent with a continuum rather than distinct categories."}
uncertainty_df <- data.frame(
  uncertainty = fit_gmm$uncertainty,
  pitch_type  = breaking$pitch_type
)

ggplot(uncertainty_df, aes(x = uncertainty, fill = pitch_type, colour = pitch_type)) +
  geom_histogram(binwidth = 0.02, alpha = 0.6, position = "identity") +
  facet_wrap(~ pitch_type, ncol = 1,
             labeller = labeller(pitch_type = label_names)) +
  scale_fill_manual(values   = label_colours, guide = "none") +
  scale_colour_manual(values = label_colours, guide = "none") +
  labs(
    x     = "Classification uncertainty (1 − max posterior probability)",
    y     = "Count",
    title = "GMM posterior classification uncertainty by pitch type"
  ) +
  theme_bw() +
  theme(panel.grid.minor  = element_blank(),
        strip.background  = element_blank())

```

```{r gmm-ari}
ari <- adjustedRandIndex(fit_gmm$classification, breaking$pitch_type)
cat(sprintf("Adjusted Rand Index (GMM clusters vs. CU/SL/ST labels): %.4f\n", ari))

tab <- table(Cluster = fit_gmm$classification, Label = breaking$pitch_type)
kable(tab, caption = sprintf(
  "GMM cluster (G = %d) × Statcast label contingency table.", fit_gmm$G))
```



**ARI interpretation:** ARI near 1 indicates clusters closely recover the Statcast
labels; ARI near 0 means the data-driven partition has little correspondence to
the CU/SL/ST taxonomy. An ARI below ~0.2 is consistent with a continuum.

---

## Stage 3: Bayesian Hierarchical Model

A Bayesian logistic model with pitcher-level random intercepts estimates whether
the Statcast pitch-type label predicts whiff probability above and beyond movement
and velocity. CU is the reference category; `beta_sl` and `beta_st` are the key
estimands. This stage addresses **predictive validity**: do the labels carry
incremental information about outcomes once physical features are controlled?

```{r jags-prep}
jags_df <- breaking |>
  dplyr::filter(!is.na(whiff)) |>
  mutate(
    pitcher_idx = as.integer(factor(pitcher_id)),
    year_idx    = as.integer(factor(game_year))
  )
set.seed(99)
jags_df <- jags_df |> slice_sample(n = 50000)
jags_data <- list(
  Y          = jags_df$whiff,
  pfx_x_std  = jags_df$pfx_x_std,
  pfx_z_std  = jags_df$pfx_z_std,
  velo_std   = jags_df$velo_std,
  spin_std   = jags_df$spin_std,
  label_sl   = jags_df$label_sl,
  label_st   = jags_df$label_st,
  pitcher_id = jags_df$pitcher_idx,
  year_id    = jags_df$year_idx,
  N          = nrow(jags_df),
  J          = max(jags_df$pitcher_idx),
  K          = n_distinct(jags_df$year_idx)
)



cat(sprintf("JAGS data: N = %s pitches, J = %s pitchers, K = %s seasons\n",
            format(jags_data$N, big.mark = ","),
            format(jags_data$J, big.mark = ","),
            format(jags_data$K, big.mark = ",")))

```

```{r jags-model-string}
model_string <- "
model {
  for (i in 1:N) {
    logit(p[i]) <- alpha[pitcher_id[i]] +
                   gamma[year_id[i]]    +
                   beta_pfx_x * pfx_x_std[i] +
                   beta_pfx_z * pfx_z_std[i] +
                   beta_velo  * velo_std[i]   +
                   beta_spin  * spin_std[i]   +
                   beta_sl    * label_sl[i]   +
                   beta_st    * label_st[i]
    Y[i] ~ dbern(p[i])
  }

  beta_st_vs_sl <- beta_st - beta_sl

  for (j in 1:J) {
    alpha[j] ~ dnorm(mu_alpha, tau_alpha)
  }

  for (k in 1:K) {
    gamma[k] ~ dnorm(0, tau_gamma)
  }

  mu_alpha  ~ dnorm(0, 0.001)
  tau_alpha ~ dgamma(0.001, 0.001)
  tau_gamma ~ dgamma(0.001, 0.001)

  beta_pfx_x ~ dnorm(0, 0.001)
  beta_pfx_z ~ dnorm(0, 0.001)
  beta_velo  ~ dnorm(0, 0.001)
  beta_spin  ~ dnorm(0, 0.001)
  beta_sl    ~ dnorm(0, 0.001)
  beta_st    ~ dnorm(0, 0.001)
}
"

model_file <- tempfile(fileext = ".jags")
writeLines(model_string, model_file)
```



```{r jags-run, cache=TRUE, results='hide'}
inits_fn <- function() list(
  beta_pfx_x = rnorm(1, 0, 0.1),
  beta_pfx_z = rnorm(1, 0, 0.1),
  beta_velo  = rnorm(1, 0, 0.1),
  beta_spin  = rnorm(1, 0, 0.1),
  beta_sl    = rnorm(1, 0, 0.1),
  beta_st    = rnorm(1, 0, 0.1),
  mu_alpha   = rnorm(1, 0, 0.1),
  tau_alpha  = rgamma(1, 1, 1),
  tau_gamma  = rgamma(1, 1, 1)
)

params <- c("beta_pfx_x", "beta_pfx_z", "beta_velo", "beta_spin",
            "beta_sl", "beta_st", "beta_st_vs_sl",
            "mu_alpha", "tau_alpha", "tau_gamma")

fit_jags <- jags(
  data               = jags_data,
  inits              = replicate(3, inits_fn(), simplify = FALSE),
  parameters.to.save = params,
  model.file         = model_file,
  n.chains           = 3,
  n.adapt            = 1000,
  n.burnin           = 5000,
  n.iter             = 10000,
  n.thin             = 2,
  parallel           = TRUE,
  verbose            = FALSE
)

```

```{r jags-convergence}
rhat_vals <- unlist(fit_jags$Rhat)
rhat_vals <- rhat_vals[!is.na(rhat_vals)]
n_bad     <- sum(rhat_vals > 1.1)
max_rhat  <- max(rhat_vals)

cat(sprintf(
  "Gelman-Rubin check — max R-hat: %.3f  |  nodes > 1.1: %d / %d  |  %s\n",
  max_rhat, n_bad, length(rhat_vals),
  if (n_bad == 0) "CONVERGED" else "NOT CONVERGED — interpret with caution"
))
```



```{r jags-posterior-table}
target_params <- c("beta_pfx_x", "beta_pfx_z", "beta_velo", "beta_spin",
                   "beta_sl", "beta_st", "beta_st_vs_sl",
                   "mu_alpha", "tau_alpha", "tau_gamma")

post_tbl <- fit_jags$summary[target_params,
                              c("mean", "sd", "2.5%", "97.5%", "Rhat"),
                              drop = FALSE] |>
  as.data.frame() |>
  rownames_to_column("Parameter") |>
  mutate(
    Parameter = recode(Parameter,
      beta_pfx_x    = "Horiz. break (β_pfx_x)",
      beta_pfx_z    = "Vert. break (β_pfx_z)",
      beta_velo     = "Velocity (β_velo)",
      beta_spin     = "Spin rate (β_spin)",
      beta_sl       = "SL vs. CU (β_sl)",
      beta_st       = "ST vs. CU (β_st)",
      beta_st_vs_sl = "ST vs. SL (β_st − β_sl, derived)",
      mu_alpha      = "Intercept hypermean (μ_α)",
      tau_alpha     = "Intercept precision (τ_α)",
      tau_gamma     = "Year RE precision (τ_γ)"
    ),
    across(where(is.numeric), ~ round(., 3))
  )

kable(
  post_tbl,
  col.names = c("Parameter", "Post. mean", "Post. SD", "2.5%", "97.5%", "R-hat"),
  caption   = paste(
    "Posterior summaries for the whiff model.",
    "Key estimands: β_sl (slider vs. curveball) and β_st (sweeper vs. curveball),",
    "controlling for horizontal break, vertical break, velocity, and spin rate.",
    "Credible intervals that include zero indicate the label adds no predictive",
    "value beyond physical features."
  )
)

```

```{r jags-posterior-plot, fig.cap="Posterior distributions for the three pairwise label contrasts. The dashed line marks zero. Distributions centred on zero with 95% CIs that include zero indicate the label does not predict whiff rate beyond physical features.", fig.height=4}
extract_samples <- function(fit, param) {
  samps <- do.call(rbind, lapply(fit$samples, as.data.frame))
  samps[[param]]
}

contrast_df <- bind_rows(
  data.frame(value    = extract_samples(fit_jags, "beta_sl"),
             contrast = "SL vs. CU"),
  data.frame(value    = extract_samples(fit_jags, "beta_st"),
             contrast = "ST vs. CU"),
  data.frame(value    = extract_samples(fit_jags, "beta_st_vs_sl"),
             contrast = "ST vs. SL (derived)")
)

contrast_colours <- c("SL vs. CU"           = "#D55E00",
                      "ST vs. CU"           = "#009E73",
                      "ST vs. SL (derived)" = "#CC79A7")

ci_df <- contrast_df |>
  group_by(contrast) |>
  summarise(lo = quantile(value, 0.025), hi = quantile(value, 0.975),
            .groups = "drop")

ggplot(contrast_df, aes(value, fill = contrast, colour = contrast)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey30") +
  geom_segment(
    data = ci_df,
    aes(x = lo, xend = hi, y = -Inf, yend = -Inf, colour = contrast),
    linewidth = 2, lineend = "round", inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  facet_wrap(~ contrast, scales = "free_y") +
  scale_fill_manual(values   = contrast_colours) +
  scale_colour_manual(values = contrast_colours) +
  labs(
    x     = expression(beta ~ "(log-odds of whiff)"),
    y     = "Posterior density",
    title = "Posterior distributions of pitch-type label effects on whiff rate",
    fill  = NULL, colour = NULL
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "none",
        strip.background = element_blank())
```


---

## Summary

```{r summary-table}
sl_ci <- round(fit_jags$summary["beta_sl",       c("2.5%", "97.5%")], 3)
st_ci <- round(fit_jags$summary["beta_st",       c("2.5%", "97.5%")], 3)
stvsl <- round(fit_jags$summary["beta_st_vs_sl", c("2.5%", "97.5%")], 3)

summary_df <- data.frame(
  Stage   = c("PCA", "GMM", "JAGS — SL vs. CU", "JAGS — ST vs. CU",
               "JAGS — ST vs. SL"),
  Finding = c(
    sprintf("PC1 explains %.1f%%; PC1+PC2 = %.1f%%",
            var_exp[1] * 100, cum_var[2] * 100),
    sprintf("Optimal G = %d (%s); ARI vs. labels = %.3f",
            fit_gmm$G, fit_gmm$modelName, ari),
    sprintf("95%% CI [%.3f, %.3f]%s", sl_ci[1], sl_ci[2],
            if (sl_ci[1] < 0 & sl_ci[2] > 0) " — includes zero" else " — excludes zero"),
    sprintf("95%% CI [%.3f, %.3f]%s", st_ci[1], st_ci[2],
            if (st_ci[1] < 0 & st_ci[2] > 0) " — includes zero" else " — excludes zero"),
    sprintf("95%% CI [%.3f, %.3f]%s", stvsl[1], stvsl[2],
            if (stvsl[1] < 0 & stvsl[2] > 0) " — includes zero" else " — excludes zero")
  ),
  Interpretation = c(
    if (cum_var[2] > 0.85) "Most variance in two dimensions — consistent with a continuum"
    else "Variance spread across PCs — possible multimodal structure",
    if (ari < 0.2) "Low ARI — clusters do not correspond to CU/SL/ST labels"
    else if (ari < 0.5) "Moderate ARI — partial label–cluster alignment"
    else "High ARI — clusters align well with labels",
    if (sl_ci[1] < 0 & sl_ci[2] > 0) "Label adds no evidence of predictive value"
    else "Label carries incremental predictive value",
    if (st_ci[1] < 0 & st_ci[2] > 0) "Label adds no evidence of predictive value"
    else "Label carries incremental predictive value",
    if (stvsl[1] < 0 & stvsl[2] > 0) "SL and ST not distinguishable on whiff outcomes"
    else "SL and ST distinguishable on whiff outcomes"
  )
)

kable(summary_df, caption = "Summary of findings across all three analysis stages.")
```

**Take-home:** The three stages together evaluate construct validity from
complementary angles — convergent (PCA), discriminant (GMM), and predictive
(JAGS). Consistent evidence across all three would be needed to conclude that
the CU/SL/ST taxonomy reflects genuine latent structure rather than a coarse
operational discretisation of a continuous movement space.
