---
title: "Breaking Ball Taxonomy: Curveball, Slider, and Sweeper"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 9
    fig_height: 5.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(mclust)
library(jagsUI)
library(httr)
library(readr)
library(knitr)
library(tibble)
library(vegan)
library(ggrepel)
```

## Data

Knit this file from the project root directory. If `data/raw/statcast_breaking.rds`
does not exist, the chunk below pulls CU, SL, and ST pitches for 2022–2025
directly from Baseball Savant (no baseballr required). This takes several minutes
and requires an internet connection.

```{r data-pull}
data_path <- file.path("data", "raw", "statcast_breaking.rds")

if (!file.exists(data_path)) {

  pull_savant_chunk <- function(start_date, end_date) {
    url <- paste0(
      "https://baseballsavant.mlb.com/statcast_search/csv?all=true",
      "&hfPT=CU%7CSL%7CST%7C",
      "&hfGT=R%7C",
      "&type=details",
      "&player_type=pitcher",
      "&game_date_gt=", start_date,
      "&game_date_lt=", end_date,
      "&min_pitches=0&min_results=0&min_pas=0"
    )
    resp <- tryCatch(
      GET(url,
          add_headers(`User-Agent` = "Mozilla/5.0 (compatible; R research)"),
          timeout(120)),
      error = function(e) NULL
    )
    if (is.null(resp) || http_error(resp)) {
      warning(sprintf("Failed: %s – %s", start_date, end_date))
      return(NULL)
    }
    txt <- content(resp, as = "text", encoding = "UTF-8")
    if (nchar(trimws(txt)) < 100 || startsWith(trimws(txt), "<")) {
      warning(sprintf("Non-CSV response: %s – %s", start_date, end_date))
      return(NULL)
    }
    tryCatch(
      read_csv(I(txt), show_col_types = FALSE, progress = FALSE),
      error = function(e) NULL
    )
  }

  # Monthly chunks for the 2022-2025 regular season (April–September).
  months <- expand.grid(
    year  = 2022:2025,
    month = 4:9,
    stringsAsFactors = FALSE
  )
  starts <- sprintf("%d-%02d-01", months$year, months$month)
  eom <- function(yr, mo) {
    as.character(as.Date(sprintf("%d-%02d-01", yr, mo %% 12 + 1)) - 1)
  }
  ends <- mapply(eom, months$year, months$month)

  raw_list <- mapply(function(s, e) {
    Sys.sleep(1)
    message(sprintf("Pulling %s to %s...", s, e))
    pull_savant_chunk(s, e)
  }, starts, ends, SIMPLIFY = FALSE)

  raw <- bind_rows(Filter(Negate(is.null), raw_list))

  dir.create(dirname(data_path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(raw, data_path)
  message(sprintf("Saved %d rows to %s", nrow(raw), data_path))

  cat(sprintf("\nTotal rows: %s\n\n", format(nrow(raw), big.mark = ",")))
  print(table(raw$pitch_type, raw$game_year))
}

breaking_raw <- readRDS(data_path)
sprintf("Loaded %s rows, %s columns", nrow(breaking_raw), ncol(breaking_raw))
```

```{r data-prep}
# Columns required downstream.
REQUIRED_COLS <- c("pitcher", "pitch_type", "p_throws",
                   "pfx_x", "pfx_z", "release_speed",
                   "release_spin_rate", "description")

breaking <- breaking_raw |>
  select(any_of(REQUIRED_COLS)) |>
  filter(pitch_type %in% c("CU", "SL", "ST")) |>
  mutate(
    # Flip pfx_x for LHP so arm-side movement is consistently positive.
    # After this: glove-side break (dominant for CU/SL/ST) is negative.
    pfx_x_adj = if_else(p_throws == "L", -pfx_x, pfx_x),

    # Outcome: per-pitch whiff indicator.
    whiff = as.integer(description == "swinging_strike"),

    # Label dummies — CU is the reference category (both = 0).
    label_sl = as.integer(pitch_type == "SL"),
    label_st = as.integer(pitch_type == "ST"),

    # Consecutive pitcher index for JAGS random effects.
    pitcher_id = as.integer(factor(pitcher))
  ) |>
  # Drop any row missing a feature used in PCA/GMM/JAGS.
  drop_na(pfx_x_adj, pfx_z, release_speed, release_spin_rate) |>
  # Standardise the four movement/speed features once for reuse downstream.
  mutate(
    pfx_x_std  = as.numeric(scale(pfx_x_adj)),
    pfx_z_std  = as.numeric(scale(pfx_z)),
    velo_std   = as.numeric(scale(release_speed)),
    spin_std   = as.numeric(scale(release_spin_rate))
  )

cat(sprintf(
  "%s pitches kept — CU: %s  SL: %s  ST: %s  |  %s unique pitchers\n",
  format(nrow(breaking), big.mark = ","),
  format(sum(breaking$pitch_type == "CU"), big.mark = ","),
  format(sum(breaking$pitch_type == "SL"), big.mark = ","),
  format(sum(breaking$pitch_type == "ST"), big.mark = ","),
  format(n_distinct(breaking$pitcher), big.mark = ",")
))
```

---

## Stage 1: PCA

PCA is run on the four standardised features — horizontal break (`pfx_x_adj`),
vertical break (`pfx_z`), velocity, and spin rate — to characterise the shape
of breaking-ball movement space. The **continuum hypothesis** predicts a single
elongated cloud in PC space rather than three distinct clusters.

```{r pca-fit}
features_scaled <- breaking |>
  select(pfx_x_std, pfx_z_std, velo_std, spin_std) |>
  as.matrix()

rda_fit  <- rda(features_scaled)
rda_smry <- summary(rda_fit)

# Site scores (one row per pitch, PC1/PC2 columns)
all_scores <- as.data.frame(scores(rda_fit, choices = 1:2)$sites)
all_scores$pitch_type <- breaking$pitch_type

# Feature loadings with Euclidean distance from origin
loadings_raw <- as.data.frame(rda_smry$species[, 1:2])
loadings_raw$feature <- rownames(loadings_raw)
loadings_raw$dist    <- sqrt(loadings_raw$PC1^2 + loadings_raw$PC2^2)
loadings_raw$label   <- recode(loadings_raw$feature,
  pfx_x_std = "H. break",
  pfx_z_std = "V. break",
  velo_std  = "Velocity",
  spin_std  = "Spin rate"
)

# Variance explained
var_exp <- rda_smry$cont$importance["Proportion Explained", ]
cum_var <- rda_smry$cont$importance["Cumulative Proportion",  ]
```

```{r pca-variance-table}
var_tbl <- data.frame(
  PC              = paste0("PC", 1:4),
  `Proportion`    = round(var_exp, 3),
  `Cumulative`    = round(cum_var, 3),
  check.names = FALSE
)
kable(var_tbl, caption = "Variance explained by each principal component.")
```

```{r pca-loadings-table}
ldg <- loadings_raw |>
  select(Feature = feature, PC1, PC2) |>
  mutate(
    Feature = recode(Feature,
      pfx_x_std = "Horiz. break (pfx_x, adj.)",
      pfx_z_std = "Vert. break (pfx_z)",
      velo_std  = "Release speed",
      spin_std  = "Spin rate"
    ),
    across(where(is.numeric), ~ round(., 3))
  )
kable(ldg, caption = "PCA species scores / loadings (vegan rda).")
```

PC1 typically captures the horizontal/vertical break trade-off: curveballs
load heavily on vertical drop; sweepers load heavily on horizontal break.
PC2 often separates velocity and spin rate from movement magnitude.

```{r pca-scatter, fig.cap="PC1 vs PC2 coloured by Statcast pitch label. Convex hulls mark each label's occupied region. An elongated, overlapping cloud supports the continuum hypothesis; non-overlapping hulls support distinct categories."}
set.seed(1)
label_colours <- c(CU = "#0072B2", SL = "#D55E00", ST = "#009E73")
label_names   <- c(CU = "Curveball", SL = "Slider",  ST = "Sweeper")

scores_plot <- all_scores |> slice_sample(n = min(n(), 25000))

hull_df <- scores_plot |>
  group_by(pitch_type) |>
  slice(chull(PC1, PC2)) |>
  ungroup()

ggplot(scores_plot, aes(x = PC1, y = PC2)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_polygon(
    data = hull_df,
    aes(group = pitch_type, colour = pitch_type, fill = pitch_type),
    alpha = 0.15
  ) +
  geom_point(aes(colour = pitch_type), alpha = 0.15, size = 0.5) +
  geom_segment(
    data = loadings_raw,
    aes(x = 0, xend = PC1, y = 0, yend = PC2),
    colour = "black", linewidth = 0.65, alpha = 0.55,
    arrow = arrow(length = unit(0.01, "npc")),
    inherit.aes = FALSE
  ) +
  geom_label_repel(
    data = loadings_raw,
    aes(x = PC1, y = PC2, label = label),
    size = 4.35, alpha = 0.8,
    label.padding = unit(0.35, "lines"),
    inherit.aes = FALSE
  ) +
  scale_colour_manual(values = label_colours, labels = label_names) +
  scale_fill_manual(values   = label_colours, labels = label_names) +
  labs(
    x      = sprintf("PC1 (%.1f%% variance)", var_exp[1] * 100),
    y      = sprintf("PC2 (%.1f%% variance)", var_exp[2] * 100),
    colour = NULL, fill = NULL,
    title  = "PCA of breaking-ball movement space (2022–2025)"
  ) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "bottom") +
  guides(
    colour = guide_legend(override.aes = list(alpha = 1, size = 2)),
    fill   = guide_legend(override.aes = list(alpha = 0.4))
  )
```

---

## Stage 2: Gaussian Mixture Modelling

GMM via `mclust` is applied to the same four scaled features. The number of
components $G$ and the covariance parameterisation are selected jointly by BIC.
Cluster–label alignment is measured with the Adjusted Rand Index (ARI).

```{r gmm-fit, cache=TRUE}
set.seed(42)
bic_obj <- mclustBIC(features_scaled, G = 2:9, verbose = FALSE)
fit_gmm <- Mclust(features_scaled, x = bic_obj, verbose = FALSE)

cat(sprintf("BIC-optimal: G = %d, model = %s\n",
            fit_gmm$G, fit_gmm$modelName))
```

```{r gmm-bic-curve, fig.cap="BIC values across candidate numbers of GMM components. Each line is a covariance parameterisation; the peak marks the preferred solution."}
bic_df <- as.data.frame(bic_obj) |>
  rownames_to_column("G") |>
  mutate(G = as.integer(G)) |>
  pivot_longer(-G, names_to = "model", values_to = "BIC") |>
  filter(!is.na(BIC))

bic_best <- bic_df |> group_by(G) |> slice_max(BIC, n = 1, with_ties = FALSE)

ggplot(bic_df, aes(G, BIC, group = model, colour = model)) +
  geom_line(alpha = 0.4, linewidth = 0.5) +
  geom_point(data = bic_best, aes(G, BIC), colour = "black", size = 2.5) +
  geom_vline(xintercept = fit_gmm$G, linetype = "dashed", colour = "red") +
  scale_x_continuous(breaks = 2:9) +
  labs(
    x      = "Number of GMM components (G)",
    y      = "BIC",
    colour = "Model",
    title  = sprintf("GMM BIC curve — optimal G = %d (%s)",
                     fit_gmm$G, fit_gmm$modelName)
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "right",
        legend.text      = element_text(size = 8))
```

```{r gmm-cluster-scatter, fig.cap="GMM cluster assignments overlaid on PC space, coloured by Statcast label. If clusters align with labels, each cluster region should be dominated by one colour."}
set.seed(1)
cluster_df <- data.frame(
  PC1        = all_scores$PC1,
  PC2        = all_scores$PC2,
  pitch_type = breaking$pitch_type,
  cluster    = factor(fit_gmm$classification)
) |>
  slice_sample(n = min(n(), 25000))

ggplot(cluster_df, aes(PC1, PC2)) +
  geom_point(aes(colour = pitch_type, shape = cluster),
             alpha = 0.2, size = 0.7) +
  scale_colour_manual(values = label_colours, labels = label_names) +
  scale_shape_manual(values = c(16, 17, 15, 3, 4, 8, 7, 13)[seq_len(fit_gmm$G)]) +
  labs(
    x      = sprintf("PC1 (%.1f%%)", var_exp[1] * 100),
    y      = sprintf("PC2 (%.1f%%)", var_exp[2] * 100),
    colour = "Statcast label",
    shape  = "GMM cluster",
    title  = sprintf("GMM clusters (G = %d) vs. Statcast labels", fit_gmm$G)
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  guides(
    colour = guide_legend(override.aes = list(alpha = 1, size = 2)),
    shape  = guide_legend(override.aes = list(alpha = 1, size = 2))
  )
```

```{r gmm-ari}
ari <- adjustedRandIndex(fit_gmm$classification, breaking$pitch_type)
cat(sprintf("Adjusted Rand Index (GMM clusters vs. CU/SL/ST labels): %.4f\n", ari))

# Contingency table
tab <- table(Cluster = fit_gmm$classification, Label = breaking$pitch_type)
kable(tab, caption = sprintf(
  "GMM cluster (G = %d) × Statcast label contingency table.", fit_gmm$G))
```

**ARI interpretation:** ARI near 1 means the clusters closely recover the
Statcast labels; ARI near 0 means the data-driven boundary has little
correspondence to the CU/SL/ST taxonomy. An ARI below ~0.2 is consistent with
a continuum.

---

## Stage 3: Bayesian Hierarchical Model

A logistic hierarchical model estimates whether the pitch-type label predicts
whiff probability above and beyond the three movement/speed features.
CU is the reference category; `beta_sl` and `beta_st` are the key estimands.

```{r jags-prep}
jags_df <- breaking |>
  filter(!is.na(whiff)) |>
  mutate(pitcher_idx = as.integer(factor(pitcher_id)))

jags_data <- list(
  Y          = jags_df$whiff,
  pfx_x_std  = jags_df$pfx_x_std,
  pfx_z_std  = jags_df$pfx_z_std,
  velo_std   = jags_df$velo_std,
  label_sl   = jags_df$label_sl,
  label_st   = jags_df$label_st,
  pitcher_id = jags_df$pitcher_idx,
  N          = nrow(jags_df),
  J          = max(jags_df$pitcher_idx)
)

cat(sprintf("JAGS data: N = %s pitches, J = %s pitchers\n",
            format(jags_data$N, big.mark = ","),
            format(jags_data$J, big.mark = ",")))
```

```{r jags-model-string}
model_string <- "
model {
  for (i in 1:N) {
    logit(p[i]) <- alpha[pitcher_id[i]] +
                   beta_pfx_x * pfx_x_std[i] +
                   beta_pfx_z * pfx_z_std[i] +
                   beta_velo  * velo_std[i]   +
                   beta_sl    * label_sl[i]   +
                   beta_st    * label_st[i]
    Y[i] ~ dbern(p[i])
  }

  beta_st_vs_sl <- beta_st - beta_sl

  for (j in 1:J) {
    alpha[j] ~ dnorm(mu_alpha, tau_alpha)
  }

  mu_alpha  ~ dnorm(0, 0.001)
  tau_alpha ~ dgamma(0.001, 0.001)

  beta_pfx_x ~ dnorm(0, 0.001)
  beta_pfx_z ~ dnorm(0, 0.001)
  beta_velo  ~ dnorm(0, 0.001)
  beta_sl    ~ dnorm(0, 0.001)
  beta_st    ~ dnorm(0, 0.001)
}
"

model_file <- tempfile(fileext = ".jags")
writeLines(model_string, model_file)
```

```{r jags-run, cache=TRUE, results='hide'}
inits_fn <- function() list(
  beta_pfx_x = rnorm(1, 0, 0.1),
  beta_pfx_z = rnorm(1, 0, 0.1),
  beta_velo  = rnorm(1, 0, 0.1),
  beta_sl    = rnorm(1, 0, 0.1),
  beta_st    = rnorm(1, 0, 0.1),
  mu_alpha   = rnorm(1, 0, 0.1),
  tau_alpha  = rgamma(1, 1, 1)
)

params <- c("beta_pfx_x", "beta_pfx_z", "beta_velo",
            "beta_sl", "beta_st", "beta_st_vs_sl",
            "mu_alpha", "tau_alpha")

fit_jags <- jags(
  data               = jags_data,
  inits              = replicate(3, inits_fn(), simplify = FALSE),
  parameters.to.save = params,
  model.file         = model_file,
  n.chains           = 3,
  n.adapt            = 1000,
  n.burnin           = 5000,
  n.iter             = 10000,
  n.thin             = 2,
  parallel           = TRUE,
  verbose            = FALSE
)
```

```{r jags-convergence}
rhat_vals <- unlist(fit_jags$Rhat)
rhat_vals <- rhat_vals[!is.na(rhat_vals)]
n_bad     <- sum(rhat_vals > 1.1)
max_rhat  <- max(rhat_vals)

cat(sprintf(
  "Gelman-Rubin check — max R-hat: %.3f  |  nodes > 1.1: %d / %d  |  %s\n",
  max_rhat, n_bad, length(rhat_vals),
  if (n_bad == 0) "CONVERGED" else "NOT CONVERGED — interpret with caution"
))
```

```{r jags-posterior-table}
target_params <- c("beta_pfx_x", "beta_pfx_z", "beta_velo",
                   "beta_sl", "beta_st", "beta_st_vs_sl",
                   "mu_alpha", "tau_alpha")

post_tbl <- fit_jags$summary[target_params,
                              c("mean", "sd", "2.5%", "97.5%", "Rhat"),
                              drop = FALSE] |>
  as.data.frame() |>
  rownames_to_column("Parameter") |>
  mutate(
    Parameter = recode(Parameter,
      beta_pfx_x    = "Horiz. break (β_pfx_x)",
      beta_pfx_z    = "Vert. break (β_pfx_z)",
      beta_velo     = "Velocity (β_velo)",
      beta_sl       = "SL vs. CU (β_sl)",
      beta_st       = "ST vs. CU (β_st)",
      beta_st_vs_sl = "ST vs. SL (β_st − β_sl, derived)",
      mu_alpha      = "Intercept hypermean (μ_α)",
      tau_alpha     = "Intercept precision (τ_α)"
    ),
    across(where(is.numeric), ~ round(., 3))
  )

kable(
  post_tbl,
  col.names = c("Parameter", "Post. mean", "Post. SD", "2.5%", "97.5%", "R-hat"),
  caption   = paste(
    "Posterior summaries for the whiff model.",
    "Key estimands: β_sl (slider vs. curveball) and β_st (sweeper vs. curveball),",
    "both controlling for movement and velocity.",
    "Credible intervals that include zero indicate the label adds no predictive",
    "value beyond movement features."
  )
)
```

```{r jags-posterior-plot, fig.cap="Posterior distributions for the three pairwise label contrasts. The dashed line marks zero. Distributions centred on zero with 95% CIs that include zero indicate the label does not predict whiff rate beyond movement.", fig.height=4}
extract_samples <- function(fit, param) {
  samps <- do.call(rbind, lapply(fit$samples, as.data.frame))
  samps[[param]]
}

contrast_df <- bind_rows(
  data.frame(value    = extract_samples(fit_jags, "beta_sl"),
             contrast = "SL vs. CU"),
  data.frame(value    = extract_samples(fit_jags, "beta_st"),
             contrast = "ST vs. CU"),
  data.frame(value    = extract_samples(fit_jags, "beta_st_vs_sl"),
             contrast = "ST vs. SL (derived)")
)

contrast_colours <- c("SL vs. CU"        = "#D55E00",
                      "ST vs. CU"        = "#009E73",
                      "ST vs. SL (derived)" = "#CC79A7")

ci_df <- contrast_df |>
  group_by(contrast) |>
  summarise(lo = quantile(value, 0.025), hi = quantile(value, 0.975),
            .groups = "drop")

ggplot(contrast_df, aes(value, fill = contrast, colour = contrast)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey30") +
  geom_segment(
    data = ci_df,
    aes(x = lo, xend = hi, y = -Inf, yend = -Inf, colour = contrast),
    linewidth = 2, lineend = "round", inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  facet_wrap(~ contrast, scales = "free_y") +
  scale_fill_manual(values   = contrast_colours) +
  scale_colour_manual(values = contrast_colours) +
  labs(
    x     = expression(beta ~ "(log-odds of whiff)"),
    y     = "Posterior density",
    title = "Posterior distributions of pitch-type label effects on whiff rate",
    fill  = NULL, colour = NULL
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "none",
        strip.background = element_blank())
```

---

## Summary

```{r summary-table}
sl_ci <- round(fit_jags$summary["beta_sl",  c("2.5%", "97.5%")], 3)
st_ci <- round(fit_jags$summary["beta_st",  c("2.5%", "97.5%")], 3)
stvsl <- round(fit_jags$summary["beta_st_vs_sl", c("2.5%", "97.5%")], 3)

summary_df <- data.frame(
  Stage   = c("PCA", "GMM", "JAGS — SL vs. CU", "JAGS — ST vs. CU",
               "JAGS — ST vs. SL"),
  Finding = c(
    sprintf("PC1 explains %.1f%%; PC1+PC2 = %.1f%%",
            var_exp[1]*100, cum_var[2]*100),
    sprintf("Optimal G = %d (%s); ARI vs. labels = %.3f",
            fit_gmm$G, fit_gmm$modelName, ari),
    sprintf("95%% CI [%.3f, %.3f]%s",
            sl_ci[1], sl_ci[2],
            if(sl_ci[1] < 0 & sl_ci[2] > 0) " — includes zero" else " — excludes zero"),
    sprintf("95%% CI [%.3f, %.3f]%s",
            st_ci[1], st_ci[2],
            if(st_ci[1] < 0 & st_ci[2] > 0) " — includes zero" else " — excludes zero"),
    sprintf("95%% CI [%.3f, %.3f]%s",
            stvsl[1], stvsl[2],
            if(stvsl[1] < 0 & stvsl[2] > 0) " — includes zero" else " — excludes zero")
  ),
  Interpretation = c(
    if(cum_var[2] > 0.85) "Most variance in two dimensions — consistent with a continuum"
    else "Variance spread across PCs — possible multimodal structure",
    if(ari < 0.2) "Low ARI — clusters do not correspond to CU/SL/ST labels"
    else if(ari < 0.5) "Moderate ARI — partial label–cluster alignment"
    else "High ARI — clusters align well with labels",
    if(sl_ci[1] < 0 & sl_ci[2] > 0) "Label adds no evidence of predictive value"
    else "Label carries incremental predictive value",
    if(st_ci[1] < 0 & st_ci[2] > 0) "Label adds no evidence of predictive value"
    else "Label carries incremental predictive value",
    if(stvsl[1] < 0 & stvsl[2] > 0) "SL and ST not distinguishable on whiff outcomes"
    else "SL and ST distinguishable on whiff outcomes"
  )
)

kable(summary_df, caption = "Summary of findings across all three analysis stages.")
```

**Take-home:** Paste this output (or a screenshot) back for interpretation.
The summary table above captures the key numbers needed to evaluate whether
the CU/SL/ST taxonomy is empirically supported.
